Overlapping Batches for the Assessment of Solution Quality in Stochastic Programs

### 2014-06-19 omrp_lb_rev.tex
corrected brackets after/inside expectations

Unless $f$ has a simple structure or the number of realizations of $\xi$ is small, (\ref{eq:sto_prog}) typically cannot be solved exactly. 
An approximation of (\ref{eq:sto_prog}) can be generated by sampling from the distribution of $\xit$.
While other sampling schemes are possible, let $\xit^1, \xit^2, \dots, \xit^m$ be an independent and identically distributed (i.i.d.) sample from the distribution of $\xit$.
A sampling approximation of (\ref{eq:sto_prog}), often called the sample average approximation, is given by
\begin{align} \tag{SP$_m$} \label{eq:sto_prog_m}
	z_m^* & = \min_{\x \in X} \frac{1}{m} \sum_{i=1}^m f(\x,\xiti).
\end{align}
We denote an optimal solution to (\ref{eq:sto_prog}) as $\xs$ and an optimal solution to (\ref{eq:sto_prog_m}) as $\xs_m$.

@@

\begin{proof}
	For brevity, we drop the $lj$ subscripts from $\db_{lj}$ and $\gb_{lj}$ and $l$ from $\gbb_l$ and $\dbb_l$.
	(i) We can decompose the expectation as
	\begin{align*}
		\e{(\db - \gb)^4} & = \e{\left[ \left.(\db - \gb)^4 \right| \db - \gb \neq 0\right]} \p{\db - \gb \neq 0} + 0\\
		& \leq \e{\left[ \left.(\db - \gb)^4 \right| \db - \gb \neq 0 \right]} \p{ \xs_{m_l}  \neq \xs}.
	\end{align*}
	The expectation term on the right-hand side is bounded by assumption A2 for i.i.d.\ samples.  
        Assumption A3 states $\p{\xs_{m_l} \neq \xs} \rightarrow 0$ exponentially fast.  
        Thus $\e{(\db - \gb)^4} \rightarrow 0$ exponentially fast.
(ii) This expectation term can be decomposed in the same manner
	\begin{align*}
		\e{(\dbb - \gbb)^4} & = \e{\left[ \left. (\dbb - \gbb)^4 \right| \dbb - \gbb \neq 0\right]} \p{\dbb - \gbb \neq 0}.
	\end{align*}
	With the expectation term on the right-hand side again bounded by A2, and the probability term converges to zero exponentially fast by Lemma \ref{lem:gbb_prob}.
\end{proof}

We are now ready to show that the resulting point estimators are strongly consistent. 
We use the abbreviation a.s.\ to denote almost sure convergence. 
We also use $\mu_{\xh} := \e{\left[f(\xh,\xit) - f(\xs,\xit)\right]}$ to denote the optimality gap of $\xh$ and $\sigma^2_{\xh} := \var{f(\xh,\xit) - f(\xs,\xit)}$ to denote the associated variance term.

@@

	The fringe batches come in identically distributed pairs.  
        The first and last batch are identically distributed, as are the second and second-to-last, etc.  
        We know this because we could reverse the order of the i.i.d.\ sample and perform the batching again to get the exact same batch structure.  
        Rewriting to take this into account, letting $A_j = (\gb_j - \gbb)^2 - (\db_j - \dbb)^2$, and continuing from (\ref{eq:vvar_conv_pf_1}),
	\begin{align*}
		& \left( \frac{n}{dt} \right)^2 n_b \e{ \sum_{j=1}^{n_b} \left[ (\gb_j - \gbb)^2 - (\db_j - \dbb)^2 \right]^2 } \\
		 =  & \left( \frac{n}{dt} \right)^2 n_b \e{\left[ \sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \left[ A_j \right]^2 \
		 + \sum_{j=\left\lceil \gammab^{-1} \right\rceil+1}^{n_b-\left\lceil \gammab^{-1} \right\rceil} \left[ A_j \right]^2 \
		 + \sum_{j=n_b-\left\lceil \gammab^{-1} \right\rceil+1}^{n_b} \left[ A_j \right]^2 \right]\
		} \\
		 = & \left( \frac{n}{dt} \right)^2 n_b \sum_{j=\left\lceil \gammab^{-1} \right\rceil+1}^{n_b-\left\lceil \gammab^{-1} \right\rceil} \e{ \left[ A_j \right]^2 } + \left( \frac{n}{dt} \right)^2 (n_b) 2\sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \e{ \left[ A_j \right]^2 } \\
		 = & \left( \frac{n}{dt} \right)^2 n_b \left(n_b - 2\left\lceil \gammab^{-1} \right\rceil \right) \e{ \left[ A_j \right]^2 }
		+ \left( \frac{n}{dt} \right)^2 (n_b) 2\sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \e{ \left[ A_j \right]^2 }.
	\end{align*}	
	We have $\left\lceil \gammab^{-1} \right\rceil + 1$ expectation terms, each of which has a coefficient that is less than $\left(\frac{n}{dt}\right)^2n_b^2 < (2m)^2$.  
        So each expectation can be bounded by
	\begin{align}
		\left( 2m \right)^2 \e{ \left[ A_j \right]^2 } & = \left( 2m \right)^2 \e{ \left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^2 \left[(\gb_j - \gbb) + (\db_j - \dbb)\right]^2 } \notag \\
		& \leq \left( 2m \right)^2 \sqrt{\e{\left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^4}} \sqrt{\e{\left[(\gb_j - \gbb) + (\db_j - \dbb)\right]^4 }} \label{eq:var_reduct_cauchy2} \\
		& \leq \left( 2m \right)^2 \sqrt{\e{\left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^4}} \sqrt{8 \left(\e{(\gb_j - \gbb)^4} + \e{(\db_j - \dbb)^4} \right)}, \label{eq:var_reduct_holder2}
	\end{align}
	where equation (\ref{eq:var_reduct_cauchy2}) follows from the Cauchy-Schwartz inequality, and equation (\ref{eq:var_reduct_holder2}) follows from H\"{o}lder's inequality.  
        Since we can bound both expectation terms in (\ref{eq:var_reduct_holder2}) by A2, we can demonstrate $\sqrt{\frac{n}{m}}(nVG - nVD) \xrightarrow{L^2} 0$ by focusing on only one of the terms.  
        Focusing on the first expectation term in (\ref{eq:var_reduct_holder2}) and returning the $\tfrac{n}{m}$, by H\"{o}lder's inequality,
	\begin{align*}
		\frac{n}{m} 4m^2 \e{\left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^4}	& \leq 32 \frac{n}{m} m^2\left( \e{(\db_j - \gb_j)^4} + \e{(\dbb - \gbb)^4} \right).
	\end{align*}	
	By Lemma \ref{lem:gb_gbb_l4}, both expectation terms above converge to zero with exponential rates.  
        With a batching structure $m_l = \lfloor n_l^r \rfloor$ for some $0<r<1$, we see that $\frac{n_l}{m_l} \e{(n_l VG_l - n_l VD_l)^2} \rightarrow 0$.
\end{proof}

>>>>

Unless $f$ has a simple structure or the number of realizations of $\xit$ is small, (\ref{eq:sto_prog}) typically cannot be solved exactly. 
An approximation of (\ref{eq:sto_prog}) can be generated by sampling from the distribution of $\xit$.
While other sampling schemes are possible, let $\xit^1, \xit^2, \dots, \xit^m$ be an independent and identically distributed (i.i.d.) sample from the distribution of $\xit$.
A sampling approximation of (\ref{eq:sto_prog}), often called the sample average approximation, is given by
\begin{align} \tag{SP$_m$} \label{eq:sto_prog_m}
	z_m^* & = \min_{\x \in X} \frac{1}{m} \sum_{i=1}^m f(\x,\xiti).
\end{align}
We denote an optimal solution to (\ref{eq:sto_prog}) as $\xs$ and an optimal solution to (\ref{eq:sto_prog_m}) as $\xs_m$.

@@

\begin{proof}
	For brevity, we drop the $lj$ subscripts from $\db_{lj}$ and $\gb_{lj}$ and $l$ from $\gbb_l$ and $\dbb_l$.
	(i) We can decompose the expectation as
	\begin{align*}
		\e{(\db - \gb)^4} & = \e{ \left.(\db - \gb)^4 \right| \db - \gb \neq 0} \p{\db - \gb \neq 0} + 0\\
		& \leq \e{ \left.(\db - \gb)^4 \right| \db - \gb \neq 0} \p{ \xs_{m_l}  \neq \xs}.
	\end{align*}
	The expectation term on the right-hand side is bounded by assumption A2 for i.i.d.\ samples.  
        Assumption A3 states $\p{\xs_{m_l} \neq \xs} \rightarrow 0$ exponentially fast.  
        Thus $\e{(\db - \gb)^4} \rightarrow 0$ exponentially fast.
(ii) This expectation term can be decomposed in the same manner
	\begin{align*}
		\e{(\dbb - \gbb)^4} & = \e{\left. (\dbb - \gbb)^4 \right| \dbb - \gbb \neq 0} \p{\dbb - \gbb \neq 0}.
	\end{align*}
	With the expectation term on the right-hand side again bounded by A2, and the probability term converges to zero exponentially fast by Lemma \ref{lem:gbb_prob}.
\end{proof}

We are now ready to show that the resulting point estimators are strongly consistent. 
We use the abbreviation a.s.\ to denote almost sure convergence. 
We also use $\mu_{\xh} := \e{f(\xh,\xit) - f(\xs,\xit)}$ to denote the optimality gap of $\xh$ and $\sigma^2_{\xh} := \var{f(\xh,\xit) - f(\xs,\xit)}$ to denote the associated variance term.

@@

	The fringe batches come in identically distributed pairs.  
        The first and last batch are identically distributed, as are the second and second-to-last, etc.  
        We know this because we could reverse the order of the i.i.d.\ sample and perform the batching again to get the exact same batch structure.  
        Rewriting to take this into account, letting $A_j = (\gb_j - \gbb)^2 - (\db_j - \dbb)^2$, and continuing from (\ref{eq:vvar_conv_pf_1}),
	\begin{align*}
		& \left( \frac{n}{dt} \right)^2 n_b \e{ \sum_{j=1}^{n_b} \left[ (\gb_j - \gbb)^2 - (\db_j - \dbb)^2 \right]^2 } \\
		 =  & \left( \frac{n}{dt} \right)^2 n_b \e{\sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \left( A_j \right)^2 \
		 + \sum_{j=\left\lceil \gammab^{-1} \right\rceil+1}^{n_b-\left\lceil \gammab^{-1} \right\rceil} \left( A_j \right)^2 \
		 + \sum_{j=n_b-\left\lceil \gammab^{-1} \right\rceil+1}^{n_b} \left( A_j \right)^2 
		} \\
		 = & \left( \frac{n}{dt} \right)^2 n_b \sum_{j=\left\lceil \gammab^{-1} \right\rceil+1}^{n_b-\left\lceil \gammab^{-1} \right\rceil} \e{ \left( A_j \right)^2 } + \left( \frac{n}{dt} \right)^2 (n_b) 2\sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \e{ \left( A_j \right)^2 } \\
		 = & \left( \frac{n}{dt} \right)^2 n_b \left(n_b - 2\left\lceil \gammab^{-1} \right\rceil \right) \e{ \left( A_j \right)^2 }
		+ \left( \frac{n}{dt} \right)^2 (n_b) 2\sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \e{ \left( A_j \right)^2 }.
	\end{align*}	
	We have $\left\lceil \gammab^{-1} \right\rceil + 1$ expectation terms, each of which has a coefficient that is less than $\left(\frac{n}{dt}\right)^2n_b^2 < (2m)^2$.  
        So each expectation can be bounded by
	\begin{align}
		\left( 2m \right)^2 \e{ \left( A_j \right)^2 } & = \left( 2m \right)^2 \e{ \left((\gb_j - \gbb) - (\db_j - \dbb)\right)^2 \left((\gb_j - \gbb) + (\db_j - \dbb)\right)^2 } \notag \\
		& \leq \left( 2m \right)^2 \sqrt{\e{\left((\gb_j - \gbb) - (\db_j - \dbb)\right)^4}} \sqrt{\e{\left((\gb_j - \gbb) + (\db_j - \dbb)\right)^4 }} \label{eq:var_reduct_cauchy2} \\
		& \leq \left( 2m \right)^2 \sqrt{\e{\left((\gb_j - \gbb) - (\db_j - \dbb)\right)^4}} \sqrt{8 \left(\e{(\gb_j - \gbb)^4} + \e{(\db_j - \dbb)^4} \right)}, \label{eq:var_reduct_holder2}
	\end{align}
	where equation (\ref{eq:var_reduct_cauchy2}) follows from the Cauchy-Schwartz inequality, and equation (\ref{eq:var_reduct_holder2}) follows from H\"{o}lder's inequality.  
        Since we can bound both expectation terms in (\ref{eq:var_reduct_holder2}) by A2, we can demonstrate $\sqrt{\frac{n}{m}}(nVG - nVD) \xrightarrow{L^2} 0$ by focusing on only one of the terms.  
        Focusing on the first expectation term in (\ref{eq:var_reduct_holder2}) and returning the $\tfrac{n}{m}$, by H\"{o}lder's inequality,
	\begin{align*}
		\frac{n}{m} 4m^2 \e{\left((\gb_j - \gbb) - (\db_j - \dbb)\right)^4}	& \leq 32 \frac{n}{m} m^2\left( \e{(\db_j - \gb_j)^4} + \e{(\dbb - \gbb)^4} \right).
	\end{align*}	
	By Lemma \ref{lem:gb_gbb_l4}, both expectation terms above converge to zero with exponential rates.  
        With a batching structure $m_l = \lfloor n_l^r \rfloor$ for some $0<r<1$, we see that $\frac{n_l}{m_l} \e{(n_l VG_l - n_l VD_l)^2} \rightarrow 0$.
\end{proof}
